Data-Driven Multi-Grid Solver for Accelerated \\ Pressure Projection

### 2022-04-08 tex/article.tex
submitted revised version

\begin{abstract}
Pressure projection is the single most computationally expensive step in an unsteady incompressible fluid simulation. \textcolor{blue}{This work demonstrates the ability of data-driven methods to accelerate the approximate solution of the Poisson equation at the heart of pressure projection.} Geometric Multi-Grid methods are identified as linear convolutional encoder-decoder networks and a data-driven smoother is developed using automatic differentiation to optimize the velocity-divergence projection. The new method is found to accelerate classic Multi-Grid methods by a factor of two to three with no loss of accuracy on eleven 2D and 3D flow cases including cases with dynamic immersed solid boundaries. The optimal parameters are found to transfer nearly 100\% effectiveness as the resolution is increased, providing a robust approach for accelerated pressure projection of unsteady flows.
\end{abstract}

@@

A key insight to accelerating pressure projection is that classic Geometric Multi-Grid (GMG) methods \cite{briggs2000multigrid} are simply \textit{linear} convolutional encoder-decoder networks, Figure \ref{fig:multigrid}. While there is a long history of optimizing multi-grid methods (both geometric \cite{tang2000sparse} and algebraic \cite{brezina2006adaptive}), few of these have taken a direct data-driven approach. \textcolor{blue}{\cite{tang2000sparse} construct a sparse approximate smoother by solving a series of least-square matrix equations instead of a data-driven approach. This improved the convergence properties on extremely skewed grids compared to Gauss-Sidel smoothers, but not on more regular grids or when using a very sparse (and therefore fast) approximate inverse.}

Employing modern data-driven techniques could  further accelerate GMG without the loss of accuracy currently limiting CNN methods.  Recent work has shown promising initial results in constructing data-driven restriction and prolongation operators \cite{KATRUTSA2020112524,greenfeld2019learning}. \textcolor{blue}{However, improving the architecture of these operators is a very challenging discrete optimization problem, limiting the results achieved thus far. Both \cite{KATRUTSA2020112524,greenfeld2019learning} show improved convergence spectra compared to classic GMG methods, but no overall speed-up. }

\textcolor{blue}{Building on this foundation, this manuscript develops a novel data-driven smoother which is optimize using automatic differentiation to minimize the velocity-divergence over the full recursive GMG V-cycle. This new framework achieves 2 to 3-fold acceleration compared to standard GMG on eleven 2D and 3D benchmark projection problems including those derived from unsteady Navier-Stokes simulations with dynamic immersed boundaries.} Moreover, since the new approach maintains linearity in the pressure and the nonlinear dependence on the matrix coefficients are scale-invariant, the data-optimized parameters generalize extremely well to new flows and simulation resolutions.

@@

The discrete Poisson equation in the projection step is defined as
\begin{equation}\label{eq:axb}
    A x = b
\end{equation}
where $x$ is the unknown pressure field vector, $b$ is the source term proportional to the divergence of the velocity field to be projected, and $A$ is the discrete matrix for the Poisson operator. For a conservative Poisson operator, $A$ is symmetric, has zero-sum rows and columns, is negative semi-definite, and has a single zero eigenvalue. \textcolor{blue}{The zero eigenvalue means the solution is unique up to an additive constant, in agreement with the derivative action of the pressure on the velocity-divergence. While it is possible to add an additional condition to the matrix, this is not required and has no impact on the projection problem or the measured pressure forces.} The matrix is also extremely sparse. While the solution vector size $N$ may easily be $10^6-10^8$, a second-order scheme on structured grids in $M$ dimensions result in only $M$ non-zero sub-diagonals.

Iterative methods solve equation \ref{eq:axb} by updating an approximate solution $x^k$ to a new solution $x^{k+1}$ with reduced error. As the problem is linear, the equation for the update $\epsilon$ is simply
\begin{equation}\label{eq:aer} 
    A \epsilon^k = r^k \equiv b - Ax^k
\end{equation}
where $r$ is the residual. \textcolor{blue}{In the context of the pressure Poisson equation, the residual is the remaining divergence in the velocity field. Driving this residual to zero by updating the pressure is the goal of the projection step.}

In practise, only an approximate solution for $\epsilon$ is obtained, after which the solution and residual are incremented
\begin{equation}\label{eq:increment}
    x^{k+1} = x^k+\epsilon^k, \quad r^{k+1} = r^k-A\epsilon^k.
\end{equation}
\textcolor{blue}{This process is iterated until the velocity-divergence residual is reduced to a specified tolerance, at which point the projection step is complete.}

@@

While more complex options are certainly possible, this work considers a sparse approximate inverse as a smoother
\begin{equation}\label{eq:smooth}
    \epsilon = \tilde A^{-1}r
\end{equation}
where $\tilde A^{-1}=f(A\,|\theta)$ is a parameterized pseudo-inverse with the same sparsity as $A$ and $\theta$ is the parameter vector. The advantage of this pseudo-inverse smoother is speed. Unlike Gauss-Sidel or SOR smoothers, the sparse matrix-vector multiplication of equation \ref{eq:smooth} requires no back-propagation and so can be vectorized in serial or parallel implementations, offering a significant potential speed-up. Additionally, the matrix $A$ is constant during the projection step (and often for an entire simulation), meaning $\tilde A^{-1}$ can be stored ahead of time, avoiding on the fly function evaluations.

@@

The parameterized smoother is optimized for a set of training data $X=\{A,r^0\}$. The loss is defined as the reduction in the log-$L_2$-norm of the residual after each V-cycle iteration 
\begin{equation}
L = \log_{10}(|r^{k+1}|_2 / |r^{k}|_2)
\end{equation}
This loss is averaged over each iteration and each example in the data-set. \textcolor{blue}{Working directly with a residual loss has a number of advantages. First and foremost, the purpose of the pressure projection step is to drive the velocity-divergence residual to zero, making this the ultimate judge of any projection method. Second, this incorporates problem-specific residual data which would be missing from a metric based only on the V-cycle operator. Finally, the residual is always available, unlike the error with respect to the unknown exact pressure solution.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/synthetic}
    \caption{\textcolor{blue}{Examples from the 2D static (left), dipole (center), and sphere (right) synthetic cases. The initial residual $r^0$ (row-1) and 10x residual after one data-driven V-cycle $10r^1$ (row-2) plotted on the same scale. The solution after one data-driven V-cycle $x^1$ (row-3) and it's 10x error compared to a machine precision solution $10\epsilon^1$ (row-4) plotted on the same scale.}}
    \label{fig:synthetic cases}
\end{figure}

@@

\textcolor{blue}{A set of six synthetic cases were developed to establish the ability of the data-driven approach to project out residuals located at (i) domain boundaries, (ii) within the fluid, or (iii) on body boundaries. The three 2D cases and their highly localize residuals are shown in Figure~\ref{fig:synthetic cases}, and each has a matching 3D case. The `static' cases feature a randomized pressure gradient throughout the domain produced by residuals only on the domain boundaries (i). The `dipole' and `sphere' cases feature residuals on a random dipole in the fluid (ii) or on the boundary of a random immersed sphere (iii). Neumann conditions are applied on the domain and sphere boundaries using the Boundary Data Immersion Method (BDIM) to adjust the $A$ matrix coefficients, as described and validated in \cite{maertens2015accurate,Lauber2022}. The  generative equations for each synthetic case are given in \ref{app synth}.}

A data-driven GMG solver is optimized for each synthetic case using 100 randomized training examples. \textcolor{blue}{Figure~\ref{fig:synthetic cases} shows the residual and error of each case after one data-driven V-cycle. The results show that the initially highly-localized residual is projected throughout the domain, without build-up of error on the domain or immersed boundaries. Repeated V-cycles drives the residual and error to machine zero in every case.}

Figure~\ref{fig:cross plot} quantifies the generalization performance of each trained projection method on 100 unseen test examples for each case. \textcolor{blue}{The results for a Jacobi-smoothed GMG V-cycle are shown for comparison as this is the initial condition for the \textit{untrained} data-driven method.} After training, the data-driven method vastly outperform the Jacobi V-cycle, doubling the residual loss on the 3D-dipole case and providing nearly 30-fold improvement on the 3D-static case. While the performance of the data-driven projection method is best when testing and training on the same case, most of the solvers still improve on the Jacobi-smoothed GMG performance for cases outside of their training set. Indeed, the solver trained on the 2D-sphere generalizes essentially as well as a solver trained on the `union' of the all training case data. The exceptions are the static cases, with the solver trained on the 2D-static case failing to converge on three of the other test cases. Note that despite identical $A$ matrices in the static and dipole cases, the performance of the those solvers are completely different. This indicates that the data-driven methods are specializing based on the residual data as well as the matrix structure, a feature unique to this approach.

@@

\textcolor{blue}{Finally, the acceleration of the data-driven projection method is evaluated against high-performance GMG solvers using classic stationary smoothers; Gauss-Sidel and SOR.} The `union'-trained data-driven solver is chosen and 100 tests are generated for each case on different grids to test the ability of the data-driven method to generalize to new grid sizes.

The results in Figure~\ref{fig:synthetic time} show that the data-driven method has no issue projecting the velocity divergence on unseen grid sizes, reducing the residual by $10^{-3}$ in only 2-5x the time of a single Jacobi-smoothed V-cycle. \textcolor{blue}{This is chosen as our unit of time since Jacobi-smoothing requires the smallest possible number of operations per V-cycle on a given grid.} As the pseudo-inverse smoother is many times faster than the Gauss-Sidel and SOR smoothers, this results in an overall 80-210\% speed-up (mean: 133\%) relative to classic GMG solvers. Note that this speed-up will be even more favorable in parallel computation as the new smoother operates directly on the local residual without back-propagation, eliminating any additional communication.

@@

Five unsteady incompressible simulation cases were used to further characterize the accelerated projection method, Figure \ref{fig:simulation cases}. \textcolor{blue}{The first three cases are variations on standard unsteady flow benchmarks, while the second two are variations of recent validated biologically-inspired flows \cite{Lauber2022}.} The simulations feature significantly different physics to test the accelerated projection performance from flows with and without background velocities, immersed geometries, body motion, and body deformation, as well as using computational grids of different sizes and aspect ratios. \textcolor{blue}{All simulations use the same extensively validated unsteady incompressible Navier-Stokes BDIM solver \cite{maertens2015accurate,Lauber2022} and 100 Poisson matrices $A$ and initial residuals $r^0$ are copied from each unsteady simulation to form the training and testing data. The details of each case are found in \ref{app flow}.}

Figure \ref{fig:simulation cases} shows the pressure solutions, initial residual field, and the residual after a single V-cycle of the data-driven GMG method trained on each case. The results show that the residual decreases throughout the domain, including at the domain and immersed boundaries. \textcolor{blue}{This is further verified by Fig~\ref{fig:simulation forces}, which shows the unsteady pressure force the swimming shark case using the data-driven GMG solver is indistinguishable from the Gauss-Sidel smoothed results.} As such, the new data-driven method is shown to produce uniformly valid solutions despite only being trained to optimize the global residual loss.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{figures/forces.pdf}
    \caption{\textcolor{blue}{Integrated thrust and side pressure forces the swimming shark test case using the Gauss-Sidel (GS) and data-driven ($\tilde A^{-1}$) GMG solvers and $|r^k/r^0|<10^{-3}$ tolerance level.}}
    \label{fig:simulation forces}
\end{figure}

@@

The potential of machine learning advances to improve fluid dynamics is vast, but well-applied classical methods and constraints are needed to focus this potential. Wherever possible, this work has made the simplest choice in parameterization, leaving significant opportunities for future improvements. \textcolor{blue}{The flow cases in this manuscript use Cartesian-grids, but this does not limit the generality of the projection problems as the resulting $A$ matrices are nonuniform due to the presence of the immersed geometries. The current data-driven GMG framework can therefore be readily extended to the nonuniform matrices induced by stretched structured grids. Extensions to unstructured grids would require the use of algebraic instead of geometric multi-grid, and a similar data-driven sparse smoother could accelerate such projection methods.}

\bibliography{mybibfile}
\color{blue}
\appendix
\section{Synthetic case details}\label{app synth}

>>>>

\begin{abstract}
Pressure projection is the single most computationally expensive step in an unsteady incompressible fluid simulation. \textcolor{black}{This work demonstrates the ability of data-driven methods to accelerate the approximate solution of the Poisson equation at the heart of pressure projection.} Geometric Multi-Grid methods are identified as linear convolutional encoder-decoder networks and a data-driven smoother is developed using automatic differentiation to optimize the velocity-divergence projection. The new method is found to accelerate classic Multi-Grid methods by a factor of two to three with no loss of accuracy on eleven 2D and 3D flow cases including cases with dynamic immersed solid boundaries. The optimal parameters are found to transfer nearly 100\% effectiveness as the resolution is increased, providing a robust approach for accelerated pressure projection of unsteady flows.
\end{abstract}

@@

A key insight to accelerating pressure projection is that classic Geometric Multi-Grid (GMG) methods \cite{briggs2000multigrid} are simply \textit{linear} convolutional encoder-decoder networks, Figure \ref{fig:multigrid}. While there is a long history of optimizing multi-grid methods (both geometric \cite{tang2000sparse} and algebraic \cite{brezina2006adaptive}), few of these have taken a direct data-driven approach. \textcolor{black}{In \cite{tang2000sparse}, a sparse approximate smoother is constructed by solving a series of least-square matrix equations instead of a data-driven approach. This improved the convergence properties on extremely skewed grids compared to Gauss-Sidel smoothers, but not on more regular grids or when using a very sparse (and therefore fast) approximate inverse.}

Employing modern data-driven techniques could  further accelerate GMG without the loss of accuracy currently limiting CNN methods.  Recent work has shown promising initial results in constructing data-driven restriction and prolongation operators \cite{KATRUTSA2020112524,greenfeld2019learning}. \textcolor{black}{However, improving the architecture of these operators is a very challenging discrete optimization problem, limiting the results achieved thus far. Both \cite{KATRUTSA2020112524,greenfeld2019learning} show improved convergence spectra compared to classic GMG methods, but no overall speed-up. }

\textcolor{black}{Building on this foundation, this manuscript develops a novel data-driven smoother which is optimized using automatic differentiation to minimize the velocity-divergence over the full recursive GMG V-cycle. This new framework achieves 2 to 3-fold acceleration compared to standard GMG on eleven 2D and 3D benchmark projection problems including those derived from unsteady Navier-Stokes simulations with dynamic immersed boundaries.} Moreover, since the new approach maintains linearity in the pressure and the nonlinear dependence on the matrix coefficients are scale-invariant, the data-optimized parameters generalize extremely well to new flows and simulation resolutions.

@@

The discrete Poisson equation in the projection step is defined as
\begin{equation}\label{eq:axb}
    A x = b
\end{equation}
where $x$ is the unknown pressure field vector, $b$ is the source term proportional to the divergence of the velocity field to be projected, and $A$ is the discrete matrix for the Poisson operator. For a conservative Poisson operator, $A$ is symmetric, has zero-sum rows and columns, is negative semi-definite, and has a single zero eigenvalue. \textcolor{black}{The zero eigenvalue means the solution is unique up to an additive constant, in agreement with the derivative action of the pressure on the velocity-divergence. While it is possible to add an additional condition to the matrix, this is not required and has no impact on the projection problem or the measured pressure forces.} The matrix is also extremely sparse. While the solution vector size $N$ may easily be $10^6-10^8$, a second-order scheme on structured grids in $M$ dimensions result in only $M$ non-zero sub-diagonals.

Iterative methods solve equation \ref{eq:axb} by updating an approximate solution $x^k$ to a new solution $x^{k+1}$ with reduced error. As the problem is linear, the equation for the update $\epsilon$ is simply
\begin{equation}\label{eq:aer} 
    A \epsilon^k = r^k \equiv b - Ax^k
\end{equation}
where $r$ is the residual. \textcolor{black}{In the context of the pressure Poisson equation, the residual is the remaining divergence in the velocity field. Driving this residual to zero by updating the pressure is the goal of the projection step.}

In practise, only an approximate solution for $\epsilon$ is obtained, after which the solution and residual are incremented
\begin{equation}\label{eq:increment}
    x^{k+1} = x^k+\epsilon^k, \quad r^{k+1} = r^k-A\epsilon^k.
\end{equation}
\textcolor{black}{This process is iterated until the velocity-divergence residual is reduced to a specified tolerance, at which point the projection step is complete.}

@@

While more complex options are certainly possible, this work considers a sparse approximate inverse as a smoother
\begin{equation}\label{eq:smooth}
    \epsilon = \tilde A^{-1}r
\end{equation}
where $\tilde A^{-1}=f(A\,|\theta)$ is a parameterized pseudo-inverse with the same sparsity as $A$ and $\theta$ is the parameter vector. The advantage of this pseudo-inverse smoother is speed. Unlike Gauss-Sidel or SOR smoothers, the sparse matrix-vector multiplication of equation \ref{eq:smooth} requires no back-propagation and so can be vectorized in serial or parallel implementations, offering a significant potential speed-up. Additionally, the matrix $A$ is constant during the projection step (and often for an entire simulation), meaning $\tilde A^{-1}$ can be computed and stored ahead of time.

@@

The parameterized smoother is optimized for a set of training data $X=\{A,r^0\}$. The loss is defined as the reduction in the log-$L_2$-norm of the residual after each V-cycle iteration 
\begin{equation}
L = \log_{10}(|r^{k+1}|_2 / |r^{k}|_2)
\end{equation}
This loss is averaged over each iteration and each example in the data-set. \textcolor{black}{Working directly with a residual loss has a number of advantages. First and foremost, the purpose of the pressure projection step is to drive the velocity-divergence residual to zero, making this the ultimate judge of any projection method. Second, this incorporates problem-specific residual data which would be missing from a metric based only on the V-cycle operator. Finally, the residual is always available, unlike the error with respect to the unknown exact pressure solution.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/synthetic}
    \caption{\textcolor{black}{Examples from the 2D static (left), dipole (center), and sphere (right) synthetic cases. The initial residual $r^0$ (row-1) and 10x residual after one data-driven V-cycle $10r^1$ (row-2) plotted on the same scale. The solution after one data-driven V-cycle $x^1$ (row-3) and it's 10x error compared to a machine precision solution $10\epsilon^1$ (row-4) plotted on the same scale.}}
    \label{fig:synthetic cases}
\end{figure}

@@

\textcolor{black}{A set of six synthetic cases were developed to establish the ability of the data-driven approach to project out residuals (i) on domain boundaries, (ii) within the fluid, or (iii) on body boundaries. The three 2D cases and their highly localized residuals are shown in Figure~\ref{fig:synthetic cases}, and each has a matching 3D case. The `static' cases feature a random linear pressure field generated by residuals on the domain boundaries (i). The `dipole' and `sphere' cases feature residuals on a random dipole in the fluid (ii) or on the boundary of a random immersed sphere (iii). Neumann conditions are applied on the domain and sphere boundaries using the Boundary Data Immersion Method (BDIM) to adjust the $A$ matrix coefficients, as described and validated in \cite{maertens2015accurate,Lauber2022}. The  generative equations for each synthetic case are given in \ref{app synth}.}

A data-driven GMG solver is optimized for each synthetic case using 100 randomized training examples. \textcolor{black}{Figure~\ref{fig:synthetic cases} also shows the residual and error of each case after one data-driven V-cycle. The results show that the initially highly-localized residual is projected throughout the domain, without build-up of error on the domain or immersed boundaries. Repeated V-cycles drives the residual and error to machine zero in every case.}

Figure~\ref{fig:cross plot} quantifies the generalization performance of each trained projection method on 100 unseen test examples across cases. \textcolor{black}{The results for a Jacobi-smoothed GMG V-cycle are shown for comparison as this is the initial condition for the \textit{untrained} data-driven method.} After training, the data-driven method vastly outperform the Jacobi V-cycle, doubling the residual loss on the 3D-dipole case and providing nearly 30-fold improvement on the 3D-static case. While the performance of the data-driven projection method is best when testing and training on the same case, most of the solvers still improve on the Jacobi-smoothed GMG performance for cases outside of their training set. Indeed, the solver trained on the 2D-sphere generalizes essentially as well as a solver trained on the `union' of the all training case data. The exceptions are the static cases, with the solver trained on the 2D-static case failing to converge on three of the other test cases. Note that despite identical $A$ matrices in the static and dipole cases, the performance of the those solvers are completely different. This indicates that the data-driven methods are specializing based on the residual data as well as the matrix structure, a feature unique to this approach.

@@

\textcolor{black}{Finally, the acceleration of the data-driven projection method is evaluated against high-performance GMG solvers using classic stationary smoothers; Gauss-Sidel (GS) and Successive-Over-Relaxation (SOR).} The `union'-trained data-driven solver is used and 100 tests are generated for each case on different grids to test the ability of the data-driven method to generalize to new grid sizes.

The results in Figure~\ref{fig:synthetic time} show that the data-driven method has no issue projecting the velocity divergence on unseen grid sizes, reducing the residual by $10^{-3}$ in only 2-5x the time of a single Jacobi-smoothed V-cycle. \textcolor{black}{This is chosen as our unit of time since Jacobi-smoothing requires the smallest possible number of operations per V-cycle, normalizing results across grids.} As the pseudo-inverse smoother is many times faster than the Gauss-Sidel and SOR smoothers, this results in an overall 80-210\% speed-up (mean: 133\%) relative to classic GMG solvers. Note that this speed-up will be even more favorable in parallel computation as the new smoother operates directly on the local residual without back-propagation, eliminating any additional communication.

@@

Five unsteady incompressible simulation cases were used to further characterize the accelerated projection method, Figure \ref{fig:simulation cases}. \textcolor{black}{The first three cases are variations on standard unsteady flow benchmarks, while the second two are variations of recent validated biologically-inspired flows \cite{Lauber2022}.} The simulations feature significantly different physics to test the accelerated projection performance from flows with and without background velocities, immersed geometries, body motion, and body deformation, as well as using computational grids of different sizes and aspect ratios. \textcolor{black}{All simulations use the same extensively validated unsteady incompressible Navier-Stokes BDIM solver \cite{maertens2015accurate,Lauber2022} and 100 Poisson matrices $A$ and initial residuals $r^0$ are copied from each unsteady simulation to form the training and testing data. The details of each case are found in \ref{app flow}.}

Figure \ref{fig:simulation cases} shows the pressure solutions, initial residual field, and the residual after a single V-cycle of the data-driven GMG method trained on each case. The results show that the residual decreases throughout the domain, including at the domain and immersed boundaries. \textcolor{black}{This is further verified by Fig~\ref{fig:simulation forces}, which demonstrates that the unsteady swimming shark pressure forces using the data-driven GMG solver is indistinguishable from the Gauss-Sidel smoothed results.} As such, the new data-driven method is shown to produce uniformly valid solutions despite only being trained to optimize the global residual loss.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{figures/forces.pdf}
    \caption{\textcolor{black}{Integrated thrust and side pressure forces the swimming shark test case using the Gauss-Sidel (GS) and data-driven ($\tilde A^{-1}$) GMG solvers and $|r^k/r^0|<10^{-3}$ tolerance level.}}
    \label{fig:simulation forces}
\end{figure}

@@

The potential of machine learning advances to improve fluid dynamics is vast, but well-applied classical methods and constraints are needed to focus this potential. Wherever possible, this work has made the simplest choice in parameterization, leaving significant opportunities for future improvements. \textcolor{black}{The flow cases in this manuscript use Cartesian-grids, but this does not limit the generality of the projection problems as the resulting $A$ matrices are nonuniform due to the presence of the immersed geometries. The current data-driven GMG framework can therefore be readily extended to the nonuniform matrices induced by stretched structured grids. Extensions to unstructured grids would require the use of algebraic instead of geometric multi-grid, and a similar data-driven sparse smoother could accelerate such projection methods.}

\bibliography{mybibfile}
\color{black}
\appendix
\section{Synthetic case details}\label{app synth}
